\documentclass{article}
\usepackage{graphicx} % Required for inserting images
\usepackage{amsmath}

\title{Geometric distribution}

\begin{document}

\maketitle

\textbf{Examples:}
\begin{enumerate}
    \item Suppose we sample students (with replacement) from campus until a student with height less than or equal to 60 inches is found. \\\\
    Since the sampling is with replacement, the result of a given trial is independent of the other trials. Therefore, this experiment can be modeled by \( X \sim \text{Geom}(p) \) distribution, where
    \[
    P(\text{Success}) = P(\text{finding a student with height less than or equal to 60 inches})
    \]
    and the event
    \[
    \{ X = k \} := \{\text{we find $k$ students before the first student with height less than or equal to 60 inches}\}
    \]
    In this setting, \( p_X(k) = P(X = k) \), is the probability of picking exactly \( k \) students before finding a student with height less than or equal to 60 inches. \\\\
    Also, \( P(X \ge k) \) represents the probability of picking more than or equal to \( k \) students before finding a student with height less than or equal to 60 inches.

    \item Suppose we sample students (with replacement) from campus until we find one that has taken a statistics course. \\\\
    This experiment can be modeled by \( X \sim \text{Geom}(p) \) distribution, where
    \[
    P(\text{Success}) = P(\text{finding a student who has taken a statistics course})
    \]
    and the event
    \[
    \{ X = k \} := \{\text{we pick $k$ students before finding the first student who has taken at least one statistics class}\}
    \]
    In this setting, \( p_X(k) = P(X = k) \), is the probability of picking exactly \( k \) students before finding a student who has taken at least one statistics class.

    \item Suppose we choose potatoes (with replacement, even though that does not really make sense) until we find a potato that does not have a blemish. \\\\
    In this setting, a potato is a success if it does not have any blemishes. \\\\
    This experiment can be modeled by \( X \sim \text{Geom}(p) \) distribution, where
    \[
    P(\text{Success}) = P(\text{finding a potato without any blemishes})
    \]
    and the event
    \[
    \{ X = k \} := \{\text{we pick $k$ potatoes before finding the first potato without any blemishes}\}
    \]
    In this setting, \( p_X(k) = P(X = k) \), is the probability of picking exactly \( k \) potatoes before finding the first potato without any blemishes.
\end{enumerate}


\end{document}
